Computing likelihoods under the SMC for a general class of ARGs.

### 2024-03-21 paper.tex
Make Peter last author, remove Alison

% First authors
\author[1, $\dagger$]{Gertjan Bisschop}

% Middle Authors
\author[2]{Alison Etheridge}
\author[3]{Peter Ralph}

% last author
\author[1]{Jerome Kelleher}
\affil[$\dagger$]{Denotes corresponding author}

@@

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022}
or parsimony~\citep{ignatieva_kwarg_2021} and that return
a fully precise ARG estimate~\citep{wong_general_2023},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023} that return an more approximate structure.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive grown in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

Model-based inference is a central element of population genetics.
The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
tajima_evolutionary_1983} has proved to be a useful model, and is the
basis of a wide variety of inferences [some citations?]. The incorporation
of recombination in to the coalescent process~\citep{hudson_properties_1983}
has proved to be much more difficult.
% FIXME rephrase, next two sentences, lifted from ARG paper
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
nielsen_estimation_2000,
fearnhead_estimating_2001}.
% TODO reuse some of the ideas from the SMC paper here, it has
% some good explanations.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The introduction of the Sequentially Markovian Coalescent (SMC)
approximation to the coalescent with recombination
by~\cite{mcvean_approximating_2005} was a seminal contribution,
and has made many different inferences
% TODO add some more recent stuff
feasible~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
In particular, the SMC is the basis for ARG sampling methods
like ARGweaver~\citep{rasmussen_genome-wide_2014}, which remains the
% TODO list other places ARGweaver has come out on top in terms of accuracy.
gold-standard in terms of accuracy~\citep{brandt2022evaluation}

\begin{itemize}
    \item Problem: we can't use models on inferred ARGs (Relate, tsinfer, ARGneedle). These methods produce point estimates. What we can't do:
        \begin{itemize}
            \item quantify uncertainty
            \item explore ARG space systematically around the point estimates. We don't have a metric on what an improvement might be.
            \item compare models of demography
            \item can't evaluate goodness of fit of model estimates
        \end{itemize}
    \item what we do here
\end{itemize}

[TL;DR]: way to think about genealogies is tied to the CwR. However,
CwR does not enable us to simply compute a likelihood in the absence of a fully
detailed history. In the past, this has given rise to approximations to the
coalescent that had similar properties yet were significantly easier to use
as a basis for inference.
A similar process has taken place in ARG-based inference,
where scalable ARG inference has been made possible due to both these approximations
to the coalescent as well as the idea of ARGs without explicit, fully detailed
recombination histories. What is left however is to connect the two.]

%% ARGs are awesome, or general intro about approximations

[FIX ME: introduction is very much work in progress.
start is a mess. Need to find the right way in.]
Knowledge of the exact genealogy of a set of samples would make certain problems
in population genetics (e.g.\ inferring recombination rates) trivial.
Recent advancements in large-scale inference methods of the intricately
interwoven paths of inheritance referred to as ancestral recombination graphs (ARGs)
\citep{rasmussen_genome-wide_2014, heine_bridging_2018,
kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023},
are revolutionising the way we look at large genomic datasets.

Quite a number of studies have started to demonstrate the value of the recent ability to
infer ARGs at scale by emphasizing the ability to rephrase many fundamental problems
in genomics as questions on the transmission of genetic material from generation
to generation [REFS].
Yet even though the potential of ARGs in population and statistical genetics has
been demonstrated, the fact that these inferred ARGs lack explicit detail on
recombination events means that it remains unclear how to connect these
reconstructed ARGs to the stochastic/generative models that have lead us
to study these datastructures in the first place.

For a very long time people have assumed that what is meant by an ARG is the
fully specified genealogical history where all events are uniquely positioned
both in time and along the genome [Wong et al. 2023].
Yet, the ease by which such a fully detailed history can be simulated under the
CwR, is not matched by the usability of the model for inference.
In particular, given an ARG, there will always be an infinite amount of other
ARGs that could have generated the observed genomic dataset
[potentially with very different likelihoods under the CwR]
\citep{mcvean_approximating_2005}.

%% approximations have driven the rate of progress in inference
Instead, approximations of the coalescence with recombination (CwR) have
driven our ability to perform statistical genomic inference. Two approximations
in particular have proven to be instrumental: the sequentially Markovian coalescent,
and its extensions (SMC) on the one hand, and the Li and Stephens model on the other.
The impact of both models is tightly linked to the fact that they have enabled
the reformulation of inference problems as hidden Markov models (HMM) and thus
the use of the well-known efficient HMM-algorithms.

To improve the ability to simulate the CwR, McVean and Cardin
provided an approximation of the CwR that is both Markovian backwards in time
as well as left-to-right along the genome \citep{mcvean_approximating_2005}.
By sacrificing precision on our ability to capture long range linkage
information, the SMC became the engine of many powerful inference methods [PSMC].
A later improvement, the SMC', closed the gap between the SMC and the
CwR even further.

The second key approximation is not a generative model. Rather than providing
a mathematical description of the genealogies underlying the observed data
the Li and Stephens or copying model pragmatically details the relationship
of a single genome to a larger set of genomes as the result of an
incomplete copying process \citep{li_modeling_2003}. This heuristic approach
has been successfully applied for demographic inference [and other things: ADD]
\citep{sheehan_estimating_2013, steinrucken_inference_2019}.

Both approaches have further formed the basis (of many) of the recent advancements in
ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018,
kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}.
A further classification of these methods could be made based on the level of detail
that is provided on recombination events. The first real breakthrough in large-scale
ARG inference, \argweaver, heads the first group.
The ARGs sampled by \argweaver from the posterior distribution
given the observed variational data, are all fully precise and detailed in the sense
that all events are explicitly inferred. Note however that the set of possible node times
has to be limited to enable the usage of an HMM. The other more recent methods, although
still SMC-based, have added heuristic pre- and/or post-processing steps
that have allowed them to extend their approach to larger sample sizes. We should also
mention \kwarg and \texttt{ARGinfer} in this category of methods producing fully detailed ARGs
although these are not based on the SMC or the copying algorithm. In fact they represent
two extremes along the approximation continuum with \kwarg being entirely
parsimony-based while \texttt{ARGinfer} is a CwR-based MCMC sampler.

The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant
leap in scalability. These methods derive their scalability at least to a certain
degree from the fact that the ARGs they infer are approximate structures. In particular,
recombination is an emergent property of the inferred ARGs, explaining tree topology
changes going from one local tree to the next. The recombination events themselves
are not explicitly inferred and are thus not part of the resulting data structure.

The computational gains from not having to pinpoint recombination events tie in
with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS].
Some approaches have embraced the approximate nature of the resulting ARG by, for
example, highlighting the inability to resolve certain splits and instead representing
them as polytomies \citep{kelleher_inferring_2019}. Taken together this means that the
degree of completeness in which genetic inheritance from ancestors to descendants is documented
by each of these methods varies extensively.
[TO ADD: these methods only provide point estimates]

%% The Big problem: how do you connect such approximate structures to a generative model

This has complicated the ability to connect the output of this diverse range of
ARG inference methods to a model-based description that would allow us to perform
probabilistic exploration of the ARG space around the point estimate,
or likelihood-based inference.

%% what we will do
Here, the idea is to bridge this gap and use the backwards-in-time formulation of the SMC
to compute likelihoods for for \textit{any} inferred ARG
by integrating out the time to recombination.
The application of the SMC in a way that resonates more
with the true direction of flow of genetic information
results in a concrete way to formulate an algorithm
that is \textit{general} in the sense that it can deal with the various levels of
completeness that characterise the current state of the field of ARG reconstruction.

@@

>>>>

% First authors
\author[1, $\dagger$]{Gertjan Bisschop}
\author[1]{Jerome Kelleher}
% last author
\author[3]{Peter Ralph}

\affil[$\dagger$]{Denotes corresponding author}

@@

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
% TODO add SINGER
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022}
or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023}.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

Perhaps the clearest route to statistical rigor is through scalable
model-based inference, a central element of population genetics.
The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
tajima_evolutionary_1983} has proved to be a useful model, and is the
basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
The incorporation
of recombination in the coalescent process~\citep{hudson_properties_1983}
has proved to be much more difficult.
% FIXME rephrase, next two sentences, lifted from ARG paper
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
nielsen_estimation_2000,
fearnhead_estimating_2001}.
% TODO reuse some of the ideas from the SMC paper here, it has
% some good explanations.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The introduction of the Sequentially Markovian Coalescent (SMC)
approximation to the coalescent with recombination
by~\cite{mcvean_approximating_2005} was a seminal contribution,
and has made many different inferences
% TODO add some more recent stuff
feasible~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
In particular, the SMC is the basis for ARG sampling methods
like ARGweaver~\citep{rasmussen_genome-wide_2014}, which remains the
% TODO list other places ARGweaver has come out on top in terms of accuracy.
gold-standard in terms of accuracy~\citep{brandt2022evaluation}

% THE PROBLEM!!! OMG it's terrible!!!
Although major breakthroughs in scale have been achieved by hueristic
ARG inference
methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
these methods are currently fundamentally incompatible with rigorous
inference under population genetic models such as the SMC. This is 
because these methods return approximate ARG
\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. Specifically,
we cannot evaluate a likelihood of these ARGs in coalescent 
models under the classical 
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present.
[NOTE the following is super weak, and needs a lot of work.]
This inability to connect the results of large-scale inference methods
with a generative model
is a fundamental limitation, and limits both the utility of 
current point estimates as well as the possibilities for further
inference. Specifically, we are unable to quantify uncertainty 
around heuristic point estimates. We cannot explore ARG space
around the point estimates using (e.g. MCMC), as we have no way definition
of improvement. We cannot compare models of demography, or evaluate the 
goodness of fit for model estimates.

%% This is how we solve it
[TODO figure out the gARG, tree sequence etc mess]
Here, we bridge this gap between models and the products of approximate
inference by defining a likelihood function for any 
``genome ARG''~\citep{wong_general_2023} under the SMC. Using a novel
formulation of the SMC that focuses on the contributions of nodes and 
edges [FIXME] in a backwards-in-time way, we provide an efficient 
algorithm that can compute a likelihood for any gARG under this model.
We demonstrate the flexibility of this approach by XXX, and show
[SOMETHING].

@@

%% NOTE: this is some orphan text from the intro that can be rejiggered for the 
% discussion. We can talk about precision of recombination estimates a bit 
% here I think, as it is an important topic.

% Both approaches have further formed the basis (of many) of the recent advancements in
% ARG inference methods \citep{rasmussen_genome-wide_2014, heine_bridging_2018,
% kelleher_inferring_2019, speidel_method_2019, rasmussen_espalier_2022, zhang_biobank-scale_2023}.
% A further classification of these methods could be made based on the level of detail
% that is provided on recombination events. The first real breakthrough in large-scale
% ARG inference, \argweaver, heads the first group.
% The ARGs sampled by \argweaver from the posterior distribution
% given the observed variational data, are all fully precise and detailed in the sense
% that all events are explicitly inferred. Note however that the set of possible node times
% has to be limited to enable the usage of an HMM. The other more recent methods, although
% still SMC-based, have added heuristic pre- and/or post-processing steps
% that have allowed them to extend their approach to larger sample sizes. We should also
% mention \kwarg and \texttt{ARGinfer} in this category of methods producing fully detailed ARGs
% although these are not based on the SMC or the copying algorithm. In fact they represent
% two extremes along the approximation continuum with \kwarg being entirely
% parsimony-based while \texttt{ARGinfer} is a CwR-based MCMC sampler.

% The second group of methods (\tsinfer, \relate, \argneedle) has marked a significant
% leap in scalability. These methods derive their scalability at least to a certain
% degree from the fact that the ARGs they infer are approximate structures. In particular,
% recombination is an emergent property of the inferred ARGs, explaining tree topology
% changes going from one local tree to the next. The recombination events themselves
% are not explicitly inferred and are thus not part of the resulting data structure.

% The computational gains from not having to pinpoint recombination events tie in
% with the inevitable uncertainty associated with ARG reconstruction [Hayman, 2022, ADD REFS].
% Some approaches have embraced the approximate nature of the resulting ARG by, for
% example, highlighting the inability to resolve certain splits and instead representing
% them as polytomies \citep{kelleher_inferring_2019}. Taken together this means that the
% degree of completeness in which genetic inheritance from ancestors to descendants is documented
% by each of these methods varies extensively.
% [TO ADD: these methods only provide point estimates]
